HealpixNet: Efficient spherical Convolutional Neural Network with Healpix sampling for cosmological applications

### 2018-07-04 scnn.tex
small edits

Following~\cite{shuman2013emerging}, the graph normalized graph Laplacian
defined as $\L = \I - \D^{-1/2} \W \D^{-1/2}$ is a second order differential operator
that can be used to define the graph Fourier basis. Here $D$ is the diagonal
matrix where $\D_{ii}=\b{d}_i$. By construction the Laplacian is symetric positive
semi-definite and hence can be decomposed as $\L=\U \bLambda \U^*$, where $U$ is an
orthonormal matrix of eigenvectors and $\bLambda$, the diagonal matrix of
eigenvalues. The graph Fourier basis is defined as the Laplacian eigenvectors.
The graph Fourier transform of $f$ is simply its projection on $U$ given by
$\hat{\b{f}}=\U^*\b{f}$. Similarly the inverse graph Fourier transform reads $\b{f}=\U\hat{\b{f}}$.
Note that instead of frequencies, the Fourier modes are crescently ordered with
respect of the Laplacian eigenvalues $\bLambda$. In a sense, the Laplacian
eigenvalues correspond to the squared frequencies.

As shown in \figref{graph_harmonics}, it turns out that using the graph
construction described above, the graph Fourier harmonics resemble the
classical spherical harmonics giving us an indication that the graph is able to
capture the spherical properties of the HealPix sampling. This topic is further discusses in Appendix~\ref{sec:spherical_harmonics}.

@@

The major difference with the classical convolution is that the convolution kernel
$h$ cannot be thought as a $n \times n$ pattern, but is a continuous function
applied to the graph eigenvalues $\bLambda$. For visualization purposes, one can
look at the effect of the convolution on a Kronecker, i.e. one column of the matrix
$h(\L)$. 
Due to the non-regularity of the graph (i.e. the fact
that there is not perfect sampling on the sphere), the convolution effect will slightly
differ from one node to another. In the specific case of the full sphere, these
differences are negligible in most of the cases. However, when considering only subpart of
the sphere, one can observe important border effects.

\paragraph{Example:}
Let us consider the heat diffusion problem
\begin{equation} \label{eqn:heat_equation}
\L \b{f}(t) = \tau \partial_t \b{f}(t),
\end{equation}
where $\b{f}(t): \mathbb{R}_+ \rightarrow \mathbb{R}^N$. Given the initial condition
$\b{f}(0)$, the solution of~\eqnref{heat_equation} can be expressed as
\begin{equation}
\b{f}(t) = e^{-\L \tau t} \b{f}(0) = \U e^{-\bLambda t \tau} \U^* \g{f}(0) = K_t(\L) \b{f}(0),
\end{equation}
which is a convolution of the signal $\b{f}(0)$ with the kernel $K_t(x)=e^{-\tau
t x}$. Here the kernel $K_t$ is applied to the graph eigenvalues $\bLambda$ that
are one generalization of the squared frequencies on the graph. Hence the kernel $K_t$ is also a generalization of the
Gaussian kernel on the sphere. In \figref{gaussian_filters_comparizon} top, we show
the effect of the convolution by diffusing a unit of heat for $\tau=1$ and various
time $t$. In \figref{gaussian_filters_visualization} of
Appendix~\ref{sec:filter_visualization}, we show different visualizations of the
convolution kernel $K_t$. We also compare the graph convolution with the
spherical symmetric Gaussian smoothing
(\figref{gaussian_filters_comparizon} bottom) and observe that both
techniques lead to similar results. While the graph convolution is different
from a true spherical convolution, it can, providing the correct
parameterization, well approximate it.

@@

\begin{figure}[!ht]
	\centering
	\includegraphics[width=0.45\textwidth]{figures/pooling.pdf}
	\caption{Two levels of coarsening and pooling: groups of 4 cells are merged into one, then the data on them is summarized in one. The coarser cell covers $1/12$ of the sphere.}
	\label{fig:pooling}
\end{figure}

@@

\begin{figure}[!ht]
\centering
\includegraphics[width=0.45\textwidth]{figures/psd_sigma3.pdf}
\caption{Power spectrum densities of the noiseless maps. The maps have be smoothed with a Gaussian kernel of radius $3$ arcmins in the preprocessing step.}
\label{fig:psd_sigma3}
\end{figure}

@@

We compare the performance of the spherical CNN to two simple benchmark
algorithms. The two baselines are based on building features with a) the Power
Spectral Densities or b) the histogram of pixels in the maps~\cite{patton2017cosmologicalconstraints}.
After a normalization, those features are then classified use a Support Vector
Machine (SVM) with a linear kernel. Other kernels did not provide better result
while having worse scaling properties. All classifiers are trained with noisy samples. For fairness, we augmented the dataset
of the baseline algorithm in a similar way as for the network, i.e, with
created an "unlimited" number of samples by adding new random realizations of
the noise. We stop adding new samples in the classifier once the validation
error converge to a value. This process is detailled in Appendix~\ref{sec:dataset_augmentation}.

@@

\subsection{Filter visualization} 
\assign{Nathanaël, Tomek, Michaël} 
Onecommon visualization for CNN is to display the shape of the learned filters.
Since our construction lead to almost (up to the sampling irregularities)
spherical filters, we plot in \figref{learned_filter} both the section (on the
sphere) and a gnomeview projection on a plane of a random selection of learned
filters. The filters are selected from the last layer of the network
corresponding the experiment with order $o=2$ for a relative noise level of
$2$. Nevertheless, all networks presented similar gnomonic and section pattern, at least from a
human perspective. The gnomonic projection clearly exibits the sampling issue
of HealPix. For a detail, on how the convolution kernels are plot, please refer to
Appendix~\ref{sec:filter_visualization}. 
\todo{Tomek: can you say something about the filters.}

@@

\label{sec:spherical_harmonics}
\begin{figure*}[!htb]
\centering
\includegraphics[width=0.95\textwidth]{figures/eigenvectors.pdf}
\caption{12 first graph Fourier eigenvectors. Eigenvectors 1-3 could be assotiated to spherical harmonics of order 1, eigenvectors 4-8 to order 2 and 9-15 to order 3. Nevertheless, the graph eigenvectors are only approximating the real spherical harmonics.}
\label{fig:graph_harmonics}
\end{figure*}

>>>>

Following~\cite{shuman2013emerging}, the normalized graph Laplacian,
defined as $\L = \I - \D^{-1/2} \W \D^{-1/2}$, is a second order differential operator
that can be used to define a Fourier basis on the graph. Here $\D$ is the diagonal
matrix where $\D_{ii}=\b{d}_i$. By construction the Laplacian is symetric positive
semi-definite and hence can be decomposed as $\L=\U \bLambda \U^*$, where $\U$ is an
orthonormal matrix of eigenvectors and $\bLambda$, the diagonal matrix of
eigenvalues. The graph Fourier basis is defined as the Laplacian eigenvectors.
The graph Fourier transform of $f$ is simply its projection on $\U$ given by
$\hat{\b{f}}=\U^*\b{f}$. Similarly, the inverse graph Fourier transform reads $\b{f}=\U\hat{\b{f}}$.
Note that instead of frequencies, the Fourier modes are ordered in the increasing order of the Laplacian eigenvalues $\bLambda$. In a sense, the Laplacian
eigenvalues correspond to the squared frequencies.

As shown in \figref{graph_harmonics}, it turns out that using the graph
construction described above, the graph Fourier harmonics resemble the
classical spherical harmonics giving us an indication that the graph is able to
capture the spherical properties of the Healpix sampling. This topic is further discussed in \ref{sec:spherical_harmonics}.

@@

The major difference with the classical convolution is that the convolution kernel
$h$ cannot be thought as a $n \times n$ pattern, but is a continuous function
applied to the graph eigenvalues $\bLambda$. For visualization purposes, one can
look at the effect of the convolution on a Kronecker delta, i.e., one column of the matrix
$h(\L)$. 
Due to the non-regularity of the graph (i.e. the fact
that there is no perfect sampling on the sphere), the convolution effect will slightly
differ from one node to another. In the specific case of the full sphere, these
differences are negligible in most of the cases. However, when considering only parts of
the sphere, one can observe important border effects.
\todo{can we show those?}

\paragraph{Example:}
Let us consider the heat diffusion problem
\begin{equation} \label{eqn:heat_equation}
\L \b{f}(t) = \tau \partial_t \b{f}(t),
\end{equation}
where $\b{f}(t): \mathbb{R}_+ \rightarrow \mathbb{R}^N$. Given the initial condition
$\b{f}(0)$, the solution of~\eqnref{heat_equation} can be expressed as
\begin{equation}
\b{f}(t) = e^{-\L \tau t} \b{f}(0) = \U e^{-\bLambda t \tau} \U^* \g{f}(0) = K_t(\L) \b{f}(0),
\end{equation}
which is a convolution of the signal $\b{f}(0)$ with the kernel $K_t(x)=e^{-\tau
t x}$. Here the kernel $K_t$ is applied to the graph eigenvalues $\bLambda$ that
are one generalization of the squared frequencies on the graph. Hence the kernel $K_t$ is also a generalization of the
Gaussian kernel on the sphere. In \figref{gaussian_filters_comparizon} top, we show
the effect of the convolution by diffusing a unit of heat for $\tau=1$ and various
time $t$. In \figref{gaussian_filters_visualization} of
\ref{sec:filter_visualization}, we show different visualizations of the
convolution kernel $K_t$. We also compare the graph convolution with the
spherical symmetric Gaussian smoothing
(\figref{gaussian_filters_comparizon} bottom) and observe that both
techniques lead to similar results. While the graph convolution is different
from a true spherical convolution, it can, providing the correct
parameterization, well approximate it.

@@

\begin{figure}[!ht]
	\centering
	\includegraphics[width=0.45\textwidth]{figures/pooling.pdf}
	\caption{Two levels of coarsening and pooling: groups of 4 cells are merged into one, then the data on them is summarized in one. The coarsest cell covers $1/12$ of the sphere.}
	\label{fig:pooling}
\end{figure}

@@

\begin{figure}[!ht]
\centering
\includegraphics[width=0.45\textwidth]{figures/psd_sigma3.pdf}
\caption{Power spectrum densities of the noiseless maps. The maps have been smoothed with a Gaussian kernel of radius $3$ arcmins to remove high frequencies. The cosmological models could otherwise have been distinguished by looking at the PSDs.}
\label{fig:psd_sigma3}
\end{figure}

@@

We compare the performance of the spherical CNN to two simple benchmark
algorithms. The two baselines are based on building features with a) the Power
Spectral Densities or b) the histogram of pixels in the maps~\cite{patton2017cosmologicalconstraints}.
After a normalization, those features are then classified use a Support Vector
Machine (SVM) with a linear kernel. Other kernels did not provide better result
while having worse scaling properties. All classifiers are trained with noisy samples. For fairness, we augmented the dataset
of the baseline algorithm in a similar way as for the network, i.e, with
created an "unlimited" number of samples by adding new random realizations of
the noise. We stop adding new samples in the classifier once the validation
error converge to a value. This process is detailed in \ref{sec:dataset_augmentation}.

@@

\subsection{Filter visualization} 
\assign{Nathanaël, Tomek, Michaël} 
Onecommon visualization for CNN is to display the shape of the learned filters.
Since our construction lead to almost (up to the sampling irregularities)
spherical filters, we plot in \figref{learned_filter} both the section (on the
sphere) and a gnomeview projection on a plane of a random selection of learned
filters. The filters are selected from the last layer of the network
corresponding the experiment with order $o=2$ for a relative noise level of
$2$. Nevertheless, all networks presented similar gnomonic and section pattern, at least from a
human perspective. The gnomonic projection clearly exibits the sampling issue
of HealPix. For a detail, on how the convolution kernels are plot, please refer to
\ref{sec:filter_visualization}. 
\todo{Tomek: can you say something about the filters.}

@@

\label{sec:spherical_harmonics}
\begin{figure*}[!htb]
\centering
\includegraphics[width=0.95\textwidth]{figures/eigenvectors.pdf}
\caption{The 16 first Fourier modes (eigenvectors of the graph Laplacian) of a graph constructed from the Healpix sampling of the sphere. Eigenvectors 1--3 could be associated to spherical harmonics of degree $\ell=1$, eigenvectors 4--8 to degree $\ell=2$, and eigenvectors 9--15 to degree $\ell=3$. Nevertheless, the graph eigenvectors are only approximating the real spherical harmonics.}
\label{fig:graph_harmonics}
\end{figure*}
